---
title: "Baltimore County 911 Calls Analysis"
author: "Data Journalism Class"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Z-Score Analysis of Baltimore County 911 Calls

## Introduction

In this analysis, we'll examine 911 call data from Baltimore County to identify unusually busy or quiet days using z-scores. By calculating the number of calls per day and analyzing their statistical distribution, we can determine which days experienced an abnormal volume of emergency calls - information that could lead to important journalistic insights about public safety patterns.

A z-score measures how many standard deviations a data point is from the mean. Generally:
- Z-scores between -2 and 2 represent typical values (about 95% of the data)
- Z-scores beyond ±2 indicate unusual values that warrant investigation
- Z-scores beyond ±3 are very unusual (only about 0.3% of normally distributed data)

## Setup and Data Loading

Let's start by loading the necessary packages and our dataset:

```{r load-packages}
library(tidyverse)  # For data manipulation and visualization
library(lubridate)  # For handling dates
library(knitr)      # For nice tables
library(scales)     # For better axis formatting
```

```{r load-data}
# Load the 911 calls data
calls_data <- read_csv("https://raw.githubusercontent.com/dwillis/datajournalismbook-maryland/refs/heads/main/data/baltco_911_calls.csv")

# Display the first few rows to understand the structure
head(calls_data)
```

## Initial Data Exploration

Let's explore the basic structure and characteristics of our dataset:

```{r explore-data}
# Get a summary of the data
glimpse(calls_data)

# Check for any missing values
colSums(is.na(calls_data))

# Check the date range in our data
min_date <- min(calls_data$date)
max_date <- max(calls_data$date)
cat("Data ranges from", min_date, "to", max_date)

# Convert date to actual Date type if it's stored as string
# This assumes date is in format "YYYY-MM-DD"
calls_data <- calls_data %>%
  mutate(date = as.Date(date))
```

## Aggregating Calls by Day

Now let's count how many 911 calls occurred on each day:

```{r aggregate-by-day}
# Count calls per day
daily_calls <- calls_data %>%
  count(date, name = "call_count")

# Display the first few days
head(daily_calls, 10) %>%
  kable()

# Basic summary statistics for daily call volume
summary(daily_calls$call_count)
```

Let's visualize the time series of daily calls:

```{r daily-calls-plot}
# Plot daily call volume
ggplot(daily_calls, aes(x = date, y = call_count)) +
  geom_line(color = "darkblue", alpha = 0.7) +
  geom_point(color = "darkblue", alpha = 0.5, size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Calls in Baltimore County",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Calculating Z-Scores for Daily Call Volume

Now, let's calculate z-scores to identify statistically unusual days:

```{r calculate-z-scores}
# Calculate z-scores for daily call counts
daily_calls <- daily_calls %>%
  mutate(
    mean_calls = mean(call_count),
    sd_calls = sd(call_count),
    z_score = (call_count - mean_calls) / sd_calls
  )

# Display the data with z-scores
daily_calls %>%
  select(date, call_count, z_score) %>%
  head(10) %>%
  kable(digits = 2)
```

## Identifying Unusual Days

Let's identify days with unusually high or low call volumes (|z-score| > 2):

```{r identify-unusual-days}
# Find days with unusual call volumes
unusual_days <- daily_calls %>%
  filter(abs(z_score) > 2) %>%
  arrange(desc(abs(z_score)))

# Display unusual days
unusual_days %>%
  select(date, call_count, z_score) %>%
  kable(digits = 2)

# Count how many days were unusually busy vs. unusually quiet
unusual_days %>%
  group_by(unusual_type = if_else(z_score > 0, "Unusually High", "Unusually Low")) %>%
  summarise(count = n()) %>%
  kable()
```

Let's visualize these unusual days in the context of the full dataset:

```{r plot-unusual-days}
# Create a clean factor for the plot legend
daily_calls <- daily_calls %>%
  mutate(call_volume_type = case_when(
    z_score > 2 ~ "Unusually High",
    z_score < -2 ~ "Unusually Low",
    TRUE ~ "Normal"
  ))

# Plot with unusual days highlighted
ggplot(daily_calls, aes(x = date, y = call_count, color = call_volume_type)) +
  geom_line(color = "gray70") +
  geom_point(size = 2) +
  scale_color_manual(values = c("Normal" = "gray50", 
                               "Unusually High" = "red", 
                               "Unusually Low" = "blue")) +
  theme_minimal() +
  labs(title = "Daily 911 Calls with Unusual Days Highlighted",
       subtitle = "Unusual days have z-scores with absolute values > 2",
       x = "Date",
       y = "Number of Calls",
       color = "Call Volume",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Day of Week Patterns

Let's also check if certain days of the week typically see more 911 calls:

```{r day-of-week-analysis}
# Add day of week to our dataset
daily_calls <- daily_calls %>%
  mutate(
    day_of_week = wday(date, label = TRUE)
  )

# Calculate average calls by day of week
dow_summary <- daily_calls %>%
  group_by(day_of_week) %>%
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count),
    min_calls = min(call_count),
    max_calls = max(call_count)
  ) %>%
  arrange(match(day_of_week, c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Display summary by day of week
dow_summary %>% kable(digits = 1)

# Plot calls by day of week
ggplot(dow_summary, aes(x = day_of_week, y = avg_calls)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_calls, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Average 911 Calls by Day of Week",
       x = "Day of Week",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 0))
```

## Accounting for Day of Week Pattern

We can also calculate z-scores that account for the day of the week pattern:

```{r day-of-week-z-scores}
# Calculate day-of-week adjusted z-scores
daily_calls <- daily_calls %>%
  group_by(day_of_week) %>%
  mutate(
    dow_mean = mean(call_count),
    dow_sd = sd(call_count),
    dow_z_score = (call_count - dow_mean) / dow_sd
  ) %>%
  ungroup()

# Find days that were unusual for their day of the week
dow_unusual_days <- daily_calls %>%
  filter(abs(dow_z_score) > 2) %>%
  arrange(desc(abs(dow_z_score)))

# Display days that were unusual for their day of the week
dow_unusual_days %>%
  select(date, day_of_week, call_count, dow_mean, dow_z_score) %>%
  kable(digits = 2)
```

Let's compare our original unusual days with those that are unusual for their day of the week:

```{r compare-unusual-methods}
# Compare the two methods
comparison <- daily_calls %>%
  mutate(
    unusual_overall = abs(z_score) > 2,
    unusual_for_dow = abs(dow_z_score) > 2
  ) %>%
  filter(unusual_overall | unusual_for_dow) %>%
  select(date, day_of_week, call_count, z_score, dow_z_score, unusual_overall, unusual_for_dow)

# Display the comparison
comparison %>% kable(digits = 2)

# Count days that appear in both lists
both_unusual <- comparison %>%
  filter(unusual_overall & unusual_for_dow) %>%
  nrow()

# Count days only unusual overall
only_unusual_overall <- comparison %>%
  filter(unusual_overall & !unusual_for_dow) %>%
  nrow()

# Count days only unusual for their day of week
only_unusual_dow <- comparison %>%
  filter(!unusual_overall & unusual_for_dow) %>%
  nrow()

cat("Days unusual by both measures:", both_unusual, "\n",
    "Days only unusual overall:", only_unusual_overall, "\n",
    "Days only unusual for their day of week:", only_unusual_dow)
```

## Temporal Patterns and Trends

Let's also look for seasonal patterns or trends over time:

```{r temporal-analysis}
# Add month and year for seasonal analysis
daily_calls <- daily_calls %>%
  mutate(
    month = month(date, label = TRUE),
    year = year(date)
  )

# Calculate average calls by month
monthly_summary <- daily_calls %>%
  group_by(month) %>%
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count)
  )

# Plot calls by month
ggplot(monthly_summary, aes(x = month, y = avg_calls, group = 1)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Average 911 Calls by Month",
       x = "Month",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data")

# Show a rolling 7-day average to better see trends
daily_calls <- daily_calls %>%
  arrange(date) %>%
  mutate(rolling_avg_7day = zoo::rollmean(call_count, k = 7, fill = NA))

# Plot with 7-day rolling average
ggplot(daily_calls, aes(x = date)) +
  geom_line(aes(y = call_count), color = "gray60", alpha = 0.7) +
  geom_line(aes(y = rolling_avg_7day), color = "blue", size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Calls with 7-Day Moving Average",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Weather Events and Holidays

For the most unusual days, we might want to check if they coincide with significant events:

```{r investigate-top-outliers}
# Get the top 10 most unusual days
top_outliers <- daily_calls %>%
  arrange(desc(abs(z_score))) %>%
  head(10) %>%
  select(date, day_of_week, call_count, z_score)

# Display these days
top_outliers %>% kable(digits = 2)
```

## Journalistic Insights

Based on our z-score analysis, here are some potential angles for reporting:

ANSWER HERE

