---
title: "Car Thefts Significance"
---

Working in pairs, follow along with the demo to learn how to use confidence intervals and t.tests to determine statistical significance. Provide answers in this notebook when prompted. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the Prince George's county crime data and generate car thefts totals by month

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv")  |> 
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. Working in pairs, discuss how you would respond if your editor asked you what you think is going on with car thefts in Prince George's County.

PUT ANSWERS HERE.

```{r}

ggplot(car_thefts, aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2025)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Add the mean, standard error and calculate a confidence interval

Let's add some stats to that process - mean and standard deviation - and calculate the standard error and 95% confidence intervals.

```{r}

theft_summary <- car_thefts |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,  # Using t-distribution for small sample
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

# Display the confidence interval
theft_summary

# Make a better chart
car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper), 
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month), 
           y = theft_summary$mean_thefts + 10, 
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)), 
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 5: Describe the confidence interval. In pairs, discuss these questions:

- What does the 95% confidence interval show here?
- What figure(s) accurately reflects the number of car thefts in a typical month?
- Propose a headline about the trend in car thefts that is focused on the present and future, but informed by the data you see above.

PUT YOUR ANSWERS HERE.


### Task 6: Is the recent trend significant?

It _looks_ like car thefts are decreasing lately. But is that change real? We'll split the monthly data into two halves, and then ask if whether there's a statistically significant difference between these periods. To do that, we'll use a t test with our alpha at 0.05.

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |> 
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

# Visualize the comparison
period_summary <- car_thefts |> 
  group_by(period) |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    ci_lower = mean_thefts - qt(0.975, df = months - 1) * se_thefts,
    ci_upper = mean_thefts + qt(0.975, df = months - 1) * se_thefts
  )

period_summary |> 
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half of Data Period",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 3)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Task 7: Reflection

Given the above chart, is the difference between the two periods statistically significant? 

How would you revise your headline from Task 5, if at all?

PUT ANSWERS HERE



When you are finished, add, commit and push your notebook and submit the URL in ELMS.
