---
title: "Diversity Index for Maryland Cities"
author: "Your Name"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this notebook, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

---

## Step 1: Load and Inspect the Data

```{r load-data}
# Load the Excel file
df <- read_excel("Wk6 A3.xlsx")

# Preview structure
glimpse(df)
```

---

## Step 2: Filter for Maryland Cities with RacePop â‰¥ 10,000

```{r filter-md-cities}
md_cities <- df %>%
  filter(State == "Maryland", RacePop >= 10000)
```

---

## Step 3: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities %>%
  mutate(
    White_pct = White / RacePop,
    Black_pct = Black / RacePop,
    AmInd_pct = AmInd / RacePop,
    Asian_pct = Asian / RacePop,
    NHPI_pct = NHPI / RacePop
  )
```

---

## Step 4: Square the Race Percentages

```{r square-race}
md_cities <- md_cities %>%
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    AmInd_pct_sq = AmInd_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```

---

## Step 5: Calculate Race Probability

```{r race-prob}
md_cities <- md_cities %>%
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + AmInd_pct_sq + Asian_pct_sq + NHPI_pct_sq
  )
```

---

## Step 6: Calculate Ethnicity Percentages and Their Squares

```{r ethnicity}
md_cities <- md_cities %>%
  mutate(
    HispTotal = HISPANIC + NOT_HISPANIC,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NOT_HISPANIC / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  )
```

---

## Step 7: Multiply Race and Ethnicity Probabilities

```{r combined-prob}
md_cities <- md_cities %>%
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```

---

## Step 8: Calculate Diversity Index

```{r diversity-index}
md_cities <- md_cities %>%
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  )
```

---

## Final Diversity Index Results

```{r final-results}
md_summary <- md_cities %>%
  select(City, Diversity_Index) %>%
  arrange(desc(Diversity_Index))

md_summary
```

---

## âœ… Solution Key

### ðŸ” Most Diverse City

```{r most-diverse}
md_summary %>% slice(1)
```

### ðŸ”» Least Diverse City

```{r least-diverse}
md_summary %>% slice(n())
```

### âœï¸ Interpretation

> The **Diversity Index** quantifies the likelihood that two randomly selected individuals in a city are of different racial or ethnic backgrounds. A score of **0** indicates no diversity (everyone is the same race/ethnicity), while a score closer to **100** indicates higher diversity.  
>
> The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

---

## ðŸ“Ž Notes

- The shaded columns in the spreadsheet are not used, as per instructions.
- The calculation excludes the multi-racial category, consistent with the USA TODAY method.
