---
title: "Home Values After Transit Line Extension"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(tidyverse)
library(scales)
```

## The News Story Challenge

Local officials have claimed that the new light rail line has increased property values in a local neighborhood.

Before the transit line opened, homes in that neighborhood sold for an average of $285,000, with a standard deviation of $32,000.

Since the extension opened, your news team collected sales data from 14 recent home sales in the neighborhood:
$297,000, $305,000, $292,000, $318,000, $290,000, $311,000, $299,000, $325,000, $302,000, $288,000, $307,000, $315,000, $320,000, $294,000

As a journalist, you need to determine: **Is there statistical evidence that the transit extension has actually increased home values?**

## Task 1: Organize your data

```{r}
# Known information about home values before the transit extension
prior_mean <- 285000  # average home value
prior_sd <- 32000     # standard deviation

# Home sale prices after the transit extension opened (14 homes)
new_prices <- c(297000, 305000, 292000, 318000, 290000, 311000, 299000, 
                325000, 302000, 288000, 307000, 315000, 320000, 294000)

# Create a journalist-friendly dataset
home_data <- tibble(
  sale_id = 1:14,
  sale_price = new_prices
)

# View the data
home_data
```

## Task 2: Calculate key statistics

```{r}
# Calculate statistics on the new home price data
new_stats <- home_data %>%
  summarise(
    mean = mean(sale_price),
    sd = sd(sale_price),
    n = n(),
    min = min(sale_price),
    max = max(sale_price)
  )

new_stats
```

## Task 3: Make a chart

```{r}
# Function to format currency values
dollar_format <- function(x) {
  paste0("$", format(x/1000, big.mark=","), "K")
}

# Create a visualization of home prices with reference to previous average
ggplot(home_data, aes(x = reorder(sale_id, sale_price), y = sale_price)) +
  geom_col(fill = "seagreen", alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 3, y = prior_mean - 8000, 
           label = "Previous Average ($285K)", hjust = 0, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = dollar_format) +
  labs(
    title = "Oak Hill Home Sales After Transit Extension",
    subtitle = "Horizontal line shows previous neighborhood average of $285,000",
    x = "Recent Home Sales",
    y = "Sale Price",
    caption = "Source: County Property Records"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

## Task 4: Perform a hypothesis test

This is where we formally test the officials' claim that the transit line has increased home values.

**Hypotheses:**
- H₀: μ ≤ $285,000 (The transit extension has not increased home values)
- H₁: μ > $285,000 (The transit extension has increased home values)

```{r}
# Perform a one-sample t-test
# Since we want to know if home values increased, we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  home_data$sale_price,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

## Step 5: Interpreting the results for your news story

```{r}
# Extract the p-value
p_value <- t_test_result$p.value

# Set our significance level
alpha <- 0.05

# Compare and make a decision
if(p_value < alpha) {
  conclusion <- "The data provides statistical evidence supporting the claim that the transit line has increased home values."
} else {
  conclusion <- "The data does not provide strong statistical evidence that the transit line has increased home values."
}

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = dollar_format(prior_mean),
  `New average` = dollar_format(mean(new_prices)),
  `Increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent change` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%"),
  `p-value` = round(p_value, 3),
  `Statistically significant?` = if_else(p_value < alpha, "Yes", "No")
)

# Display the key statistics
story_stats
```

## Conclusion

```{r, echo=FALSE}
# Calculate the 95% confidence interval for the mean increase
ci <- t.test(home_data$sale_price)$conf.int
```

